Verification of Stencil Computations with Spatial Specifications

### 2016-07-07 paper.tex
sten def

\usepackage[]{hyperref}
\hypersetup{
    unicode=false,          % non-Latin characters in Acrobat's bookmarks
    pdftoolbar=true,        % show Acrobat toolbar?
    pdfmenubar=true,        % show Acrobat menu?
    pdffitwindow=false,      % page fit to window when opened
    pdftitle={},    % title
    pdfauthor={}
    pdfsubject={},   % subject of the document
    pdfnewwindow=true,      % links in new window
    pdfkeywords={keywords}, % list of keywords
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=darkblue,          % color of internal links
    citecolor=darkblue,        % color of links to bibliography
    filecolor=green,      % color of file links
    urlcolor=blue,          % color of external links
}

@@

\begin{abstract}
  Verifying the correctness of numerical computations relies on first
  giving a specification of their behaviour. In many cases, a
  numerical computation is derived from an underlying mathematical
  model, but this is often difficult to use as a specification as it
  is highly decoupled from the implementation. We have been studying
  specification techniques in the middle ground between low-level and
  high-level specifications for numerical code. In this paper, we
  focus on \emph{stencil computations}---a common idiom in numerical
  code, but one which is error-prone due to fine-grained indexing
  errors. We hypothesise that in practice, stencil computations tend
  to have a regular shape which is amenable to static analysis and
  simple specification. We describe a abstract spatial specification
  language for stencil computations that can be embedded as
  annotations in source code, along with inference, checking, and
  specification synthesis procedures. We evaluate our
  language against a corpus of numerical Fortran code, and show that
  the vast majority of stencil computations indeed have a simple,
  regular, static shape. Around 10\% of the stencils found evoke a
  programming pattern for which indexing errors are plausible. We
  investigate two stencils in detail and find that we are able to
  detect the vast majority of potential errors. This paper details our
  stencil specification language, its static semantics, inference,
  checking, and synthesis procedures, our implementation, and
  our evaluation study.
\end{abstract}

@@

In practice, most stencil computations have a regular shape that can
be described simply and abstractly with a small set of coarse-grained
spatial descriptions. We introduce a simple specification language of
such descriptions, abstracting over the fine-grained detail of stencil
access patterns. For our first Laplace example, our
inference procedure provides the following specification to line 3:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This describes that \texttt{a} is accessed with a
symmetrical pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Fig.~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Fig.~\ref{ref:navier-stokes-fragment}(c) illustrates
the specification pictorially.

In this paper, we make the following contributions:
%
\begin{itemize}[leftmargin=0.8em]
\item we introduce a specification language for
stencil computation that captures many common forms
of data access pattern (Section~\ref{sec:lang}), providing
 an equational theory and denotational model (Sect.~\ref{sec:semantics});

\item we detail inference, checking, and specification synthesis 
algorithms for our language (Sect.~\ref{sec:analysis}), leveraging the semantic model;

\item we provide an implementation of our approach as an extension to
  CamFort, an open-source program analysis tool for Fortran programs.

\item we show that around 10\% of stencils are in a class that
  represent a plausible scenario for a programming error;

\item we provide a more detailed verification case study of two
  particular stencil computation and show that we
  are able to detect the vast majority of (synthesised) possible
  indexing errors.
\end{itemize}
%

@@

\begin{definition}[Neighbourhood index]
  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is an
  integer constant. That is, a neighbourhood index is a
  relative index defined as constant translations of base induction
  variables. (The relation $\equiv$ here identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ \eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
  the predicate \neigh{}. %An array subscript $v(\bar{e})$ where every
% $e \in \bar{e}$ is a neighbourhood index is called a
% \emph{neighbourhood subscript}.
\label{def:neighbour}
\end{definition}

@@

Indexing schemes give an abstract
representation of index expressions in the source language.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours allowed under the specification.
The relationship between indexing schemes and source language
terms is given precisely by the partial function, \textsf{schematic}, which
maps syntactic array-index terms to indexing schemes. This
operation is used by the inference
and the checking procedures.
\end{definition}
%
\begin{definition}%[Schematic]
\label{def:schematic}
The partial function $\textsf{schematic}$ maps
source indexing terms $(e_1, .., e_n)$
to $n$-dimensional index schemes, defined:
%
\begin{align*}
\mathsf{schematic}(e_1, \ldots, e_n) & =
[\downarrow\!e_1 \ldots \downarrow\!e_n ]
\quad \textit{iff} \;\; \forall i . \downarrow\!e_i \neq \bot
\end{align*}
%
The $\downarrow$ operator is partial
 $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
mapping individual indexing expressions to 
an indexing scheme component, defined:
%%
\begin{align*}
\downarrow\!e
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
in an expression, and $\neigh{}$ classifies expressions which
are constant offsets from an induction variable
(Defn.~\ref{def:neighbour}). In the third case,
if the expression is not a neighbourhood index but is not
a relative index either (\ie{}, not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\downarrow\!e$ is undefined (returns $\bot$). The $\mathsf{schematic}$ function
is undefined if $\downarrow$ of any of its components $e_i$ is
undefined.

For example, assuming $i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
 = \vthreeh{-1}{\infty}{0}$. However, $\textsf{schematic}(i*2, j)$, for
 example, is
 undefined as its first index is a non-neighbour relative index, which
 is outside the scope of our specification language.
\end{definition}

@@

\begin{example}
The following three array indexing terms
$(i, j+1)$, $(1, j+1)$ and $(3, j)$ 
are given by \textsf{schematic} the 
index schemes $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
and $\vtwoh{\infty}{0}$ respectively. The following
relationship of generalises then holds:
%\[
%\xymatrix@R=0.2em@C=2em{
%[\infty \;\; 0] \ar@{-}[dr]^{\preceq_n} &  \\
%& [0 \;\; 1]  \\
%[\infty \;\; 1]  \ar@{-}[ur]_{\preceq_n} &
%}
%\]
%%
$\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1}$
and
$\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1}$. 
%%
\end{example}
\noindent
Our terminology of \emph{indexing schemes} and
\emph{generalisation} is by analogy
with the \emph{type schemes} and \emph{type generalisation} of
polymorphic types in ML~\cite{milner1978theory}. Our notion is
much more rigid however, as our schemes do not bind names
and thus generalisation does not require substitution; $\infty$ acts
as a positional placeholder.

@@

There are a variety of reasons why we might not infer a actual specification from a potential
stencil computation: \\
\noindent 
1) \textbf{Non-subset induction variables} occur when the
induction variables on the RHS are not a subset of those in the LHS. These
cases are not stencils by our definition. The degenerate case of this
is to have only constant indices on the LHS. We see lots of examples
of this in loops as accumulators \eg{} computing the sum over an array; \\
\noindent
2) \textbf{Derived induction variables} where the
indexing variable (\mintinline{fortran}{x}) is derived from an
induction variable (\mintinline{fortran}{i}) as in
\mintinline{fortran}{x = len - i};  \\
\noindent
3) \textbf{Inconsistent induction dimensions} occur when
an induction variable is used to specify more than one array dimension
on the RHS or multiple induction variables are used for the same
dimension on the RHS. These are common in matrix operations such as
LU-decomposition with assignments such as
\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.

>>>>

\usepackage[draft]{hyperref}
\hypersetup{
    unicode=false,          % non-Latin characters in Acrobat's bookmarks
    pdftoolbar=true,        % show Acrobat toolbar?
    pdfmenubar=true,        % show Acrobat menu?
    pdffitwindow=false,      % page fit to window when opened
    pdftitle={},    % title
    pdfauthor={}
    pdfsubject={},   % subject of the document
    pdfnewwindow=true,      % links in new window
    pdfkeywords={keywords}, % list of keywords
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=darkblue,          % color of internal links
    citecolor=darkblue,        % color of links to bibliography
    filecolor=green,      % color of file links
    urlcolor=blue,          % color of external links
}

@@

\begin{abstract}
  Verifying the correctness of numerical computations relies on first
  giving a specification of their behaviour. In many cases, a
  numerical computation is derived from an underlying mathematical
  model, but this is often infeasible as a specification since it
  is highly decoupled from the implementation. In this paper, we consider a 
  lightweight specification technique between low-level and
  high-level specifications for numerical code. We
  target \emph{stencil computations}---a common idiom in numerical
  code, but one which is error-prone due to fine-grained indexing
  errors. We show that in practice, stencil computations tend
  to have a regular shape which is amenable to static analysis and
  simple specification. We describe an abstract spatial specification
  language for stencil computations that can be embedded as
  annotations in source code, along with inference, checking, and
  specification synthesis. We evaluate our
  language against a corpus of numerical Fortran code, and show that
  the vast majority of stencil computations indeed have a simple,
  regular, static shape. Around 10\% of the stencils found evoke a
  programming pattern for which indexing errors are plausible. 
  We simulate relevant programming errors in two stencil examples, 
  and show that we are able to detect the vast majority of potential
  errors. This paper details our stencil specification language, a
  denotational model, inference,
  checking, and synthesis procedures, our implementation, and
  our evaluation studies.
\end{abstract}

@@

In practice, most stencil computations have a regular shape that can
be described simply and abstractly with a small set of coarse-grained
spatial descriptions. We introduce a simple specification language of
such descriptions, abstracting over the fine-grained detail of stencil
access patterns. For the Laplace example, our
inference procedure provides the following specification to line 3:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This describes that \texttt{a} is accessed with a
symmetrical pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Fig.~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Fig.~\ref{ref:navier-stokes-fragment}(c) illustrates
the specification pictorially.

In this paper, we make the following contributions:
%
\begin{itemize}[leftmargin=0.8em]
\item we introduce a specification language for stencils 
  that captures many common forms of data access pattern
  (Section~\ref{sec:lang});

\item we detail inference, checking, and specification synthesis 
algorithms for our language (Sect.~\ref{sec:analysis}), derived from
the a denotational model for the language (Sect.~\ref{sec:semantics});

\item we provide an implementation of our approach as an extension to
  CamFort, an open-source program analysis tool for Fortran;

\item we give a quantiative study of
  stencil computations in a corpus of numerical Fortran code,
  totalling one million lines of code. Our tool identifies and
  synthesises specifications for 35,000 stencil computations in the
  corpus. Approximately 10\% of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors;

\item we give a more detailed verification case study of two
  particular stencil computation and show that we
  are able to detect the vast majority of (synthesised) possible
  indexing errors.
\end{itemize}
%

@@

\begin{definition}[Neighbourhood index]
  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is an
  integer constant. That is, a neighbourhood index is a
  relative index defined as constant translations of base induction
  variables. (The relation $\equiv$ here identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ \eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices 
  by the predicate $\neigh_I$ where $I$ is a set of induction
  variables. We omit $I$ when implicit from the context.
\label{def:neighbour}
\end{definition}

@@

Indexing schemes give an abstract
representation of index expressions in the source language.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours allowed under the specification.
The relationship between indexing schemes and source language
terms is given precisely by the partial function, \textsf{schematic}, which
maps syntactic array-index terms to indexing schemes. This
operation is used by the inference
and the checking procedures.
\end{definition}
%
\begin{definition}%[Schematic]
\label{def:schematic}
The partial function $\textsf{schematic}_I$ (parameterised by a set of
induction variables $I$) maps source indexing terms $(e_1, .., e_n)$
to $n$-dimensional index schemes, defined:
%
\begin{align*}
\mathsf{schematic}_I(e_1, ..., e_n) & =
[\mathsf{s}_I(e_1) ... \mathsf{s}_I(e_n) ]
\quad \textit{iff} \;\; \forall x . \mathsf{s}_I(e_x) \neq \bot
\end{align*}
%
where $\mathsf{s}_I$ is partial and 
maps individual indexing expressions to 
 indexing scheme components in $\domainVal{}$ defined:
%%
\begin{align*}
\mathsf{s}_I(e)
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \neigh_I(e) \\
-a & e \equiv i - a \; \wedge \; \neigh_I(e) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
in an expression, and $\neigh_I$ classifies expressions which
are constant offsets from an induction variable $i \in I$
(Defn.~\ref{def:neighbour}). In the third case,
if the expression is not a neighbourhood index but is not
a relative index either (\ie{}, not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\mathsf{s}_I(e)$ is undefined (returns $\bot$). The $\mathsf{schematic}$ function
is undefined if $\mathsf{s}_I$ of any of its components $e_i$ is
undefined. We omit the parameter set of induction variables $I$ to
$\mathsf{schematic}$ when clear from the context. 
\end{definition}

@@

\begin{example}
The following three array indexing terms
$(i, j+1)$, $(1, j+1)$ and $(3, j)$ 
are given by \textsf{schematic} the 
index schemes $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
and $\vtwoh{\infty}{0}$ respectively. The following
relationship of generalises then holds:
%\[
%\xymatrix@R=0.2em@C=2em{
%[\infty \;\; 0] \ar@{-}[dr]^{\preceq_n} &  \\
%& [0 \;\; 1]  \\
%[\infty \;\; 1]  \ar@{-}[ur]_{\preceq_n} &
%}
%\]
%%
$\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1}$
and
$\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1}$.

As an example of undefined \textsf{schematic}, 
 $\textsf{schematic}(i*2, j) = \bot$ as its first index is a non-neighbour relative index, which
 is outside the scope of our specification language.
\end{example}
\noindent
Our terminology of \emph{indexing schemes} and
\emph{generalisation} is by analogy
with the \emph{type schemes} and \emph{type generalisation} of
polymorphic types in ML~\cite{milner1978theory}. Our notion is
much more rigid however, as our schemes do not bind names
and thus generalisation does not require substitution; $\infty$ acts
as a positional placeholder.

From \textsf{schematic} we now give a
formal characterisation of stencil computations for our purposes:
%%
\begin{definition}[Stencil computations]
  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
  collection of arrays of arbitrary (possible differing)
  dimensionalities.  
  A \emph{stencil computation} comprises an iteration by a set of at
  most $n$ induction variables $I$ 
  over a subset of the index space of $a$. Elements of $a$ are determined by an
  assignment $a(\bar{e}) = e_r$ where $\bar{e}$ is an array index such
  that  $\mathsf{schematic}_I(\bar{e}) \neq \bot$ and for all array
  subscripts $b(\bar{e}') \in e_r$ then
  $\mathsf{schematic}_I(\bar{e}') \neq \bot$. 
\label{def:stencil}
\end{definition}
%%

@@

There are a variety of reasons why we might not infer a actual specification from a potential
stencil computation: \\
\noindent 
1) \textbf{Non-subset induction variables} occur when the
induction variables on the RHS are not a subset of those in the LHS. These
cases are not stencils by our definition (Defn.~\ref{def:stencil},p.~\pageref{def:stencil}). The degenerate case of this
is to have only constant indices on the LHS. We see lots of examples
of this in loops as accumulators \eg{} computing the sum over an array; \\
\noindent
2) \textbf{Derived induction variables} where the
indexing variable (\mintinline{fortran}{x}) is derived from an
induction variable (\mintinline{fortran}{i}) as in
\mintinline{fortran}{x = len - i};  \\
\noindent
3) \textbf{Inconsistent induction dimensions} occur when
an induction variable is used to specify more than one array dimension
on the RHS or multiple induction variables are used for the same
dimension on the RHS. These are common in matrix operations such as
LU-decomposition with assignments such as
\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
